Sexual selection, environmental variation, range size and speciation in passerine

### 2018-10-12 Manuscript.Rmd
Luke edits 2 - intro finished

Selection pressures are the main engine of speciation. As populations diverge they accumulate genetic and phenotypic changes that ultimately erect barriers against gene flow [@Presgraves_2010; *but see* @Shaw_2011]. In this view, populations will speciate and adapt more readily to new and different environments if selection is strong and/or persistant. Hence, the rate at which speciation occurs is likely dependent on both the population's response to selection (adaptation) and the potential for divergent phenotypic traits to evolve [*e.g.* morphology or colour @Hugall_2012; @Rabosky_morph_2013]. Sexual selection affects a diverse range of phenotypic traits important to population growth and persistance [@Cally_in_review], such as reproductive success or viability, as well as the evolution of ornaments involved in mating displays or weapons used in male-male competition [@McCullough_2016]. Sexual selection is therefore hypothesised to be a key driver of diversification, via its effects on adaptation, speciation and extinction.

Sexual selection results from competition for matings, fertilisations, or associated resources, and is typically stronger in males than females [@Fromhage_2016; @Janicke_2016; @Kokko_Jennions2008]. Sexual selection has both positive and negative consequences for 'population fitness', i.e. the average absolute fitness of individuals in the population [e.g. @Kokko_Brooks_2003; @Whitlock_Agrawal_2009; @Holman_Kokko_2013]. For example, sexual selection can elevate population fitness by helping to purge deleterious mutations [@Agrawal_2001; @Siller_2001] and fix beneficial ones [@Whitlock_2000], which should mitigate the risk of extinction due to mutation accumulation or a failure to adapte to changing environments. However, sexual selection does not always favour the evolution of traits that benefit the population...

this positive correlation is not ubiquitous; indeed sexual selection frequently favours traits having negative pleiotropic effects on female fitness [@Berger_2014; @Bonduriansky_Chenoweth_2009; @Pennell_2013; @Pischedda_2006]. For instance, in species with high levels of sexual selection, mating success becomes more variable and uncertain; thus individuals must invest heavily in costly weaponry and sexual signals to increase their chances of successfully reproducing. Such high investment, may trade off against strategies that would improve offspring fitness (*e.g.* parental care). This is a costly consequence of sexual selection that affects persistence, adaptation and extinction [@Kokko_Brooks_2003].

Alongside changes to fitness, sexual selection has been proposed to facilitate speciation through the evolution of divergent signals/ornaments associated with improved mating success [@Lande_1981; @Lande_1982]. For instance, in antbirds (Thamnophilidae), genera with complex songs and striking dichromatism, are more species rich [@Seddon_2008]. Under this model of speciation, divergent sexual signals in one sex (usually males) co-evolve with divergent preferences for those signals in the opposite sex, leading to behavioural reproductive isolation [@Safran_2013]. Concurrently, sexual selection can increase the extent to which populations diverge by enlarging the phenotypic space diverging species can evolve into. These predictions emerge as sexual selection may increase genetic diversity as a result of trade-offs between naturally- and sexually-selected traits or trade-offs under intra-locus sexual conflict between male and female trait expressions [@Lorch_2003; @Bonduriansky_2011; @Radwan_2016]. However, the genetic architecture (*per se* heritability) of fitness and sexually selected traits may vary due to different sex determination systems across taxa. In male homogametic species (ZZ/ZW or ZZ/ZO) --- such as birds --- males are predicted to have increased heritability of male-specific genotypes on the Z chromosome associated with mating success: such as elaborate ornamentation or plumage colouration [@Hastings_1994; @Reeve_2003]. If fitness and associated mating traits are more heritable in birds we might expect bird populations experiencing stronger sexual selection to have increased adaptation and speciation rates compared to male heterogametic taxa.

Importantly, the effects of sexual selection on adaptation and speciation may depend on environmental conditions. Both theoretical and empirical studies suggest that under stressful/changing environments, sexual selection may have greater fitness benefits than under benign environments. Stressful environments strengthen the positive correlation between male mating success and female fitness, reducing the burden of negative pleiotropic effects [@Berger_2014; @Connallon_Hall_2016; @Long_2012; @Martinossi-Allibert_Savković_2017]. Conversely, benign or stable environments ensure consistent selection, preferentially eroding genetic variation at sexually concordant loci (*i.e.* loci where the optimum genotype is equivelant for both sexes). In this situation, genetic variation remains disproportionately at sexually antagonistic loci; causing displacement of male and female traits from their optimum. It follows from these predictions that populations with stronger sexual selection would have increased fitness and adaptive benefits, allowing them to adapt to novel environments [*reviewed in* @Candolin_2008]. In certain circumstances, sexual selection will reinforce locally adaptated phenotypes through mate choice on phenotypes that effectively advertise quality in a specific ecological context [@Boughman_2002; @Maan_2011]. Additionally, if mate choice operates on both sexes --- or there is a correlation in trait values between a pair --- assortative mating is likely to speed-up the evolution of divergent mating signals [@Lande_1981]. In several bird taxa, beak morphology is a trait that arose under natural selection and is now a trait under sexual selection [a.k.a. "magic traits"; @Servedio_2011]. In Darwin's finches (*Geospiza fortis*) divergent beak morphology is an adaptation to local food availability that has been maintained through assortative mating [@Huber_2007]. Therefore, strong sexual selection and strong natural selection --- such as environmental change over space and time --- may work synergestically to increase speciation rates.

@@

Birds are notorious for the variety of ornaments that have evolved in response to sexual selection, such as plumage colour, size dimorphism, complex courtship rituals and male song [@Gontard_1999]. The association between sexual selection and speciation has been investigated previously in birds using macroevolutionary approaches. A meta-analysis of 64 effect sizes from 20 studies found a small, but significant positive association between sexual selection and speciation across a wide range of taxa [@Kraaijeveld_2011]. Within birds, they found a positive association but much lower than that for insects and fish. However, there was large variation in effect sizes due to variation in the measures of sexual selection used and diversification proxies (*e.g.* species/subspecies richness *vs* rate estimates). Contrary to this, @Huang_2014 found no association between sexual dichromatism and speciation using tip-rate estimates from a genetic-only phylogeny [@Jetz_2012] and sexual dichromatism data from a limited spectrophotometry dataset [*n* = 918; @Armenta_2008]; similarly @Cooney_2017 found no effect of sexual dichromatism on diversification rates when using 1,306 pairs of species and human estimates of dichromatism. Given methodological advancements in tip-rate estimates [@Rabosky_diversification_2018], the association between sexual selection and speciation is worth revisiting in order to overcome previous limitations. Firstly, the strength of sexual selection in macroevolutionary studies is often estimated using measures of sexual dimorphism or dichromatism; but these are often partial measures of sexual selection and may not account for the true variation in sexual selection across taxa. Secondly, until recently, studies on the association between sexual selection and speciation have largely ignored the contribution and interaction with environmental variables to explain variation in speciation rates. Thirdly, many studies use incomplete taxon sampling, with family-level analysis restricted to well sampled avian clades [@Seddon_2008; @Huang_2014].Fourthly, phylogenetic uncertainty in the avian tree [@Jetz_2012; *but see* @Rubolini_2015] and uncertainty in tip-rate measures of speciation such as the diversification rate statistic ($\lambda_{DR}$) or node density ($\lambda_{ND}$) [@Rabosky_diversification_2018] as well as speciation and extinction tip-rates from models using BAMM (Bayesian Analysis of Macroevolutionary Mixtures) [*see*, @Beaulieu_2015; @Rabosky_Extinction_2016; @Moore_2016; @Rabosky_BAMM_flawed_2017] continue to plague efforts to identify drivers of speciation and extinction.

Here, we perform a macroevolutionary analysis of the relationship between sexual selection and speciation or extinction in passerine birds (Order: Passeriformes), arguably the largest group for which there exists comprehensive data. We build on these previous studies by using published datasets of sexual dichromatism measured from images [*n* = 5,812; @Dale_2015] alongside smaller datasets of sexual dichromatism measured from spectrophotometry [*n* = 581; @Armenta_2008] and a composite measure of male-biased sexual selection [*n* = 2,465; @Dale_2015]. We assess the relationship between sexual selection, environmental variability and their interaction with three tip-rate measures of speciation and one measure of extinction ($\lambda_{DR}$, $\lambda_{ND}$, $\lambda_{BAMM}$ and $\mu_{BAMM}$). Using expert range maps and bioclimatic variables we are able to investigate whether the effect of sexual selection on speciation is dependent on range size, seasonal variation in temperature, spatial temperature variation (across a species’ range), long-term variation in temperature between now and the last inter-glacial (LIG) as well net primary productivity (NPP).

>>>>

Selection pressures are the main engine of speciation. As populations diverge they accumulate genetic and phenotypic changes that ultimately erect barriers against gene flow [@Presgraves_2010; *but see* @Shaw_2011]. In this view, populations will speciate and adapt more readily to new and different environments if selection is strong and/or persistant. Hence, the rate at which speciation occurs is likely dependent on both the population's response to selection (adaptation) and the potential for divergent phenotypic traits to evolve [*e.g.* morphology or colour @Hugall_2012; @Rabosky_morph_2013]. Experiments suggest that sexual selection can influence the evolution of a surprisingly diverse set of traits [@Cally_in_review], and may therefore strongly influence extant patterns of phylogenetic diversity. **I removed the bit about weapons as it kinda came out of nowhere**

Sexual selection results from competition for matings, fertilisations, or associated resources, and is typically stronger in males than females [@Fromhage_2016; @Janicke_2016; @Kokko_Jennions2008]. This competition can have both positive and negative consequences for 'population fitness', i.e. the average absolute fitness of individuals in the population [e.g. @Kokko_Brooks_2003; @Whitlock_Agrawal_2009; @Holman_Kokko_2013]. For example, sexual selection can elevate population fitness by helping to purge deleterious mutations [@Agrawal_2001; @Siller_2001] and fix beneficial ones [@Whitlock_2000], which should mitigate the risk of extinction due to mutation accumulation or a failure to adapt quickly to environmental change. However, sexual selection might also promote extinction, for example by selecting for male traits that have deterimental pleiotropic effects on female traits [@Berger_2014; @Bonduriansky_Chenoweth_2009; @Pennell_2013; @Pischedda_2006], or selecting for male phenotypes with improved mating success but which harm population productivity, e.g. by harming females or under-investing in parental care **ref here - Kokko/Rankin maybe**.

** I think this next paragraph is quite muddled and needs work!**
**Also I think you might be using 'per se' wrong, you seem to mean e.g. here**
As well as affecting population demography, sexual selection has been proposed to facilitate speciation through the evolution of divergent signals/ornaments associated with improved mating success [@Lande_1981; @Lande_1982]. For instance, in antbirds (Thamnophilidae), genera with complex songs and striking dichromatism, are more species rich [@Seddon_2008]. Under this model of speciation, divergent sexual signals in one sex (usually males) co-evolve with divergent preferences for those signals in the opposite sex, leading to behavioural reproductive isolation [@Safran_2013]. Concurrently, sexual selection can increase the extent to which populations diverge by enlarging the phenotypic space diverging species can evolve into. These predictions emerge as sexual selection may increase genetic diversity as a result of trade-offs between naturally- and sexually-selected traits or trade-offs under intra-locus sexual conflict between male and female trait expressions [@Lorch_2003; @Bonduriansky_2011; @Radwan_2016]. However, the genetic architecture (*per se* heritability) of fitness and sexually selected traits may vary due to different sex determination systems across taxa. In male homogametic species (ZZ/ZW or ZZ/ZO **Is there such a thing as ZO?**) --- such as birds --- males are predicted to have increased heritability of male-specific genotypes on the Z chromosome associated with mating success: such as elaborate ornamentation or plumage colouration [@Hastings_1994; @Reeve_2003]. If fitness and associated mating traits are more heritable in birds we might expect bird populations experiencing stronger sexual selection to have increased adaptation and speciation rates compared to male heterogametic taxa.

**Not sure if this next para is needed. You already talk about SS helping adaptation, so maybe merge your favourite parts of this with that part? Also I think the value of including all those environmental variables is to control for extraneous variation in speciation rate, and therefore get a better estimate of the importance of SS as distinct from that background**
Importantly, the effects of sexual selection on adaptation and speciation may depend on environmental conditions. Both theoretical and empirical studies suggest that under stressful/changing environments, sexual selection may have greater fitness benefits than under benign environments. Stressful environments strengthen the positive correlation between male mating success and female fitness, reducing the burden of negative pleiotropic effects [@Berger_2014; @Connallon_Hall_2016; @Long_2012; @Martinossi-Allibert_Savković_2017]. Conversely, benign or stable environments ensure consistent selection, preferentially eroding genetic variation at sexually concordant loci (*i.e.* loci where the optimum genotype is equivelant for both sexes). In this situation, genetic variation remains disproportionately at sexually antagonistic loci; causing displacement of male and female traits from their optimum. It follows from these predictions that populations with stronger sexual selection would have increased fitness and adaptive benefits, allowing them to adapt to novel environments [*reviewed in* @Candolin_2008]. In certain circumstances, sexual selection will reinforce locally adaptated phenotypes through mate choice on phenotypes that effectively advertise quality in a specific ecological context [@Boughman_2002; @Maan_2011]. Additionally, if mate choice operates on both sexes --- or there is a correlation in trait values between a pair --- assortative mating is likely to speed-up the evolution of divergent mating signals [@Lande_1981]. In several bird taxa, beak morphology is a trait that arose under natural selection and is now a trait under sexual selection [a.k.a. "magic traits"; @Servedio_2011]. In Darwin's finches (*Geospiza fortis*) divergent beak morphology is an adaptation to local food availability that has been maintained through assortative mating [@Huber_2007]. Therefore, strong sexual selection and strong natural selection --- such as environmental change over space and time --- may work synergestically to increase speciation rates.

@@

As a speciose and well-characterised group of organisms with remarkable and variable sexually-selected traits such as song and colourful plumge, birds are a popular focus for macroevolutionary studies of sexual selection and diversification **refs**. A 2011 meta-analysis, covering 20 studies of birds and other taxa, found a small but significant positive association between sexual selection and speciation overall, though the average effect size was weaker in birds than in insects and fish [@Kraaijeveld_2011]. The meta-analysis also found large variation in effect sizes between studies, likely because of differences in the metrics used as estimates of diversification and the strength of sexual selection. More recently, @Huang_2014 found no association between sexual dichromatism and speciation (*n* = 918 species), as measured used spectrophotometry **OF REAL BIRDS?** [@Armenta_2008] and tip-rate estimates from a molecular-only phylogeny [@Jetz_2012]. Similarly, @Cooney_2017 found no effect of sexual dichromatism on diversification across 1,306 pairs of species, using dichromatism scores from human observers.

Here, we investigate the association between sexual selection and diversification in birds while extending previous work in multiple ways. Firstly, many macroevolutionary studies estimate the strength of sexual selection using proxies such as sexual dimorphism or dichromatism, but these are imperfect correlates. We used a very large (*n* = 5,812) dataset of sexual dichromatism measures (measured spectrophotometrically from images; @Dale_2015]), as well as a smaller but still substantial dataset (*n* = 2,465) that gives a composite measure of male-biased sexual selection that captures variation in **TRAIT NAMES HERE**. These datasets allow us to cover the great majority of bird genera, in contrast to previous work that has often focused on particular clades (*e.g.* [@Seddon_2008; @Huang_2014]). Secondly, we use the latest methods for estimating speciation and exinction rates at the tips of phylogenetic trees, including BAMM (Bayesian Analysis of Macroevolutionary Mixtures) [*see*, @Beaulieu_2015; @Rabosky_Extinction_2016; @Moore_2016; @Rabosky_BAMM_flawed_2017], as well as older but less controversial methods such as **XXX**. We also conduct analyses that allow phylogenetic uncertainty in the avian tree [@Jetz_2012; *but see* @Rubolini_2015] to affect the precision of our estimates, rather than implicitly assuming that the single best-supported phylogenetic tree is correct. Thirdly, our analysis includes several potentially important environmental variables as predictors: range size, seasonal variation in temperature, spatial temperature variation (across a species’ range), long-term variation in temperature between now and the last inter-glacial (LIG) as well net primary productivity (NPP). By including these environmental variables in our models, we statistically account for their effects on variation in diversification rates and thereby improve our estimates of the effects of sexual selection. We also test whether the effects of sexual selection are contingent on the environment, as predicted by some theories (e.g. the theory that sexual selection is more likely to elevate popualtion fitness in changeable relative to static environments; **REFS**)
